---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

# gtxlsx

<!-- badges: start -->
[![R-CMD-check](https://github.com/JanMarvin/gtxlsx/actions/workflows/check-standard.yaml/badge.svg)](https://github.com/JanMarvin/gtxlsx/actions/workflows/check-standard.yaml)
[![codecov](https://codecov.io/gh/JanMarvin/gtxlsx/graph/badge.svg?token=JMMBBBHPTG)](https://app.codecov.io/gh/JanMarvin/gtxlsx)
[![r-universe](https://janmarvin.r-universe.dev/badges/gtxlsx)](https://janmarvin.r-universe.dev/gtxlsx)

<!-- badges: end -->

Write **gt** tables — and plain HTML tables — into an **openxlsx2** workbook.

`gt` renders beautiful tables for the web. `openxlsx2` writes real
spreadsheets. `gtxlsx` is the bit in between: it takes the table you already
built and lays it out as cells, keeping the heading, the column spanners,
the row groups, the stub, the summary rows, the footnotes and the styling
you set with `tab_style()` and `tab_options()`.

Numbers stay numbers. If gt shows `$1,234.50`, the cell holds `1234.5` with
the number format `"$"#,##0.00` attached, so the sheet is still something
you can compute with.

`gt` is a suggestion rather than a dependency. If all you want is
`wb_add_html()`, `openxlsx2` is the only thing you need.

## Maintenance

I am probably the wrong person to be building this. I do not use `gt` myself,
so I am likely to miss changes on that side and to have no feel for what its
users actually need. This would be in better hands with someone who works with
`gt` day to day, and I would happily hand it over.

## Installation

You can install development versions via r-universe:
```r
install.packages(
  "gtxlsx",
  repos = c("https://janmarvin.r-universe.dev", "https://cloud.r-project.org")
)
```

Or from GitHub directly:

```r
remotes::install_github("JanMarvin/gtxlsx")
```

## A gt table

```{r}
library(gt)
library(openxlsx2)
library(gtxlsx)

tbl <- gt(head(gtcars[, c("mfr", "model", "year", "msrp")])) |>
  tab_header(title = md("The Cars of **gtcars**")) |>
  fmt_currency(columns = msrp, decimals = 0) |>
  tab_style(
    style = list(cell_fill(color = "lightblue"), cell_text(weight = "bold")),
    locations = cells_body(columns = msrp, rows = msrp > 100000)
  )

wb <- wb_workbook()$add_worksheet(grid_lines = FALSE)
wb <- wb_add_gt(wb, tbl, dims = "B2")

if (interactive()) wb$open()
```

The title lands as a merged, bold, larger row; `**gtcars**` keeps its bold
run inside the cell; the expensive cars get the blue fill; and `msrp` is a
numeric column formatted as currency.

## An HTML table

Anything that produces an HTML `<table>` can go the same way. Inline
`<style>` rules and `style=` attributes are read, `colspan` and `rowspan`
become merged ranges, and markup inside a cell becomes rich text.

```{r}
library(openxlsx2)
library(gtxlsx)

html <- '
<style>
  th { background-color: #204060; color: white; }
  td.num { text-align: right; }
</style>
<table>
  <tr><th>Account</th><th>Change</th></tr>
  <tr><td>Cash</td><td class="num">1,204.50</td></tr>
  <tr><td><b>Net</b></td><td class="num">-2,705.50</td></tr>
</table>'

wb <- wb_workbook()$add_worksheet()
wb <- wb_add_html(wb, html, dims = "A1")

if (interactive()) wb$open()
```

## What is covered

Everything `gt` applies before rendering comes through, because `gtxlsx` reads
the table gt has already built: every `fmt_*()` and `sub_*()`, the
`cols_merge_*()` family, `text_transform()`, `data_color()`,
`summary_rows()`, spanners, footnote marks, `opt_stylize()` and the rest of
`tab_options()`.

What cannot survive a spreadsheet does not: anything gt draws as a picture.
`fmt_image()` and `cols_nanoplot()` leave the cell empty, `fmt_icon()` and
`fmt_flag()` fall back to their label text, and `fmt_url()` keeps the link
text but not the link.

`wb_to_gt()` goes the other way, from a worksheet range back to a `gt`
object. It is a development toy rather than a finished feature — see its
help page before relying on it.

## See also

`inst/examples/` holds three scripts: a gallery of gt tables covering
summaries, colours, merged columns, nested spanners and theming; a set of
HTML dialects from several table packages; and a coverage probe that runs
most of the `gt` API through the writer.
